---
output:
  pdf_document: default
  html_document: default
---

```{r}
# Loading relevant libraries
library(readxl)
library(tidyverse)
library(TTR) 
library(forecast)
```

```{r}
# Importing the death data from excel
uk_death <- read_excel('Vital statistics in the UK.xlsx', sheet = 3, skip = 5)
```

```{r}
# Checking the head of the data frame
head(uk_death)
```

```{r}
# Selecting columns needed (Year,Number of deaths: United Kingdom)
death_uk <- uk_death %>% 
  select(Year,`Number of deaths: United Kingdom`)
```

```{r}
# Checking the first 6 entries.
head(death_uk)
```


```{r}
# Checking the last 6 entries.
tail(death_uk)
```
Notice that some observations are ':'

```{r}
# Checking the structure of the data frame
str(death_uk)
```
from the above, the 'Number of deaths: United Kingdom' column is stored as 'chr'

```{r}
# Cleaning and preparing the data for time series analysis.
death_uk <- death_uk %>% 
  rename(no_of_deaths =`Number of deaths: United Kingdom`) %>% 
  filter(no_of_deaths != ':') %>% 
  arrange(Year) %>% 
  select(no_of_deaths)
```

```{r}
# Converting the data to a time series 
death_uk$no_of_deaths <- as.integer(death_uk$no_of_deaths)
death_uk_ts = ts(death_uk, frequency = 1, start = 1887)
death_uk_ts
```

```{r}
# Plotting the initial number of deaths from year 1887 - 2021
plot.ts(death_uk_ts, main='Time series of number of deaths in UK (1887 -2021)')
```
The time series appears non seasonal and can probably be described using an additive model.
Time series is non seasonal, but has trend and irregular components.

```{r}
# Estimating trend by smoothing using a simple moving average of order 10.
plot.ts(SMA(death_uk_ts, n=10), 
        main='Time series showing trend of number of deaths in UK',
        ylab = 'no of deaths (SMA)')
```

TIME SERIES MODELLING

MODEL 1 -FORECASTING USING SMOOTHING
The time series can be described by an additive model, it has trend with no seasonality, therefore:
We can use Holt's Exponential Smoothing.

```{r}
# Fitting a predictive model using Holt-Winters
death_uk_ts_forcast <- HoltWinters(death_uk_ts, gamma = FALSE)
death_uk_ts_forcast
```
An alpha value approximately 0.48, is just right in the middle of 0 and 1, which means that 48% of the weight is given to the most recent observation when estimating the level. 
A beta value of 0.15 means more weight (85%) is given to the previous trend estimate (not the most recent).

```{r}
# plotting both observed and fitted data from HoltWinters forecast.
plot(death_uk_ts_forcast)
```


```{r}
# forecasting the next 10 years.
death_uk_ts_forcast2 <- forecast(death_uk_ts_forcast, h=10)
plot(death_uk_ts_forcast2)
```

The forecast in blue.
The purple area is the 80% prediction interval
The grey area is the 95% prediction interval

```{r}
# forcasted data with the 80% and 85% intervals.
death_uk_ts_forcast2
```

```{r}
# Sum of square error
death_uk_ts_forcast$SSE
```

```{r}
# Root Mean Square Error for Holt-Winters
RMSE_HW = sqrt(mean(death_uk_ts_forcast2$residuals^2, na.rm = TRUE))
RMSE_HW
```
```{r}
# Mean absolute percentage error MAPE for Holt-Winters
MAPE_HW = mean((abs(death_uk_ts_forcast2$residuals/death_uk_ts)*100), na.rm=TRUE)
MAPE_HW
```


```{r}
# ACF and Ljung box test
acf(death_uk_ts_forcast2$residuals, lag.max=20 , na.action = na.pass) 
Box.test(death_uk_ts_forcast2$residuals, lag=20, type="Ljung-Box")
```
The P-value for the LJung-test is 0.055, there is little evidence of non-zero auto correlations in the in-sample forecast errors at lags 1-20. 

```{r}
#  plotting the forecast errors to check for constant variance
plot.ts(death_uk_ts_forcast2$residuals, main='Holt-Winters Forecast Errors',
        ylab = 'Residuals')
```

```{r}
# function to plot forecast errors and overlay normal distributed data
plotForecastErrors <- function(forecasterrors) 
{ 
# make a histogram of the forecast errors: 
mybinsize <- IQR(forecasterrors)/4  
mysd <- sd(forecasterrors) 
mymin <- min(forecasterrors) - mysd*5  
mymax <- max(forecasterrors) + mysd*3 
# generate normally distributed data with mean 0 and standard deviation mysd 
mynorm <- rnorm(10000, mean=0, sd=mysd) 
mymin2 <- min(mynorm) 
mymax2 <- max(mynorm) 
if (mymin2 < mymin) { mymin <- mymin2 } 
if (mymax2 > mymax) { mymax <- mymax2 } 
# make a red histogram of the forecast errors, with the normally distributed data overlaid: 
mybins <- seq(mymin, mymax, mybinsize) 
hist(forecasterrors, col="red", freq=FALSE, breaks=mybins) 
# freq=FALSE ensures the area under the histogram = 1 
# generate normally distributed data with mean 0 and standard deviation mysd 
myhist <- hist(mynorm, plot=FALSE, breaks=mybins) 
# plot the normal curve as a blue line on top of the histogram of forecast errors: 
points(myhist$mids, myhist$density, type="l", col="blue", lwd=2) 
} 
```

```{r}
# removing NA values from the residuals
death_uk_ts_forcast2$residuals <- death_uk_ts_forcast2$residuals[!is.na(death_uk_ts_forcast2$residuals)]
```

```{r}
# plotting if the forecast errors to check if normally distributed with mean zero
plotForecastErrors(death_uk_ts_forcast2$residuals)
```

```{r}
# library to import Augmented Dickey-Fuller Test
library(tseries)
```

```{r}
# Augmented Dickey-Fuller Test
adf.test(death_uk_ts)
```
Test if series is stationary, P-value is greater than 0.05, therefore we fail to reject null hypothesis.

```{r}
# Differencing the time series to make it stationary
death_uk_ts_diff1 <- diff(death_uk_ts, differences = 1)

# Plotting the series with difference 1.
plot(death_uk_ts_diff1, main='Time series of number of deaths in UK (DIFF 1)')
```

The plot appears stationary in mean
```{r}
# Augmented Dickey-Fuller Test for difference 1
adf.test(death_uk_ts_diff1)
```
Test if series is stationary, P-value is less than 0.05, therefore we reject null hypothesis.
Difference 1 is stationary.



SELECTING ARIMA MODEL.
```{r}
# Plotting the correlogram for diff1
acf(death_uk_ts_diff1, lag.max = 20)
acf(death_uk_ts_diff1, lag.max = 20, plot = FALSE)
```
from the correlogram, the autocorrelation at lag 1 (-0.446) exceeds the significance bounds. so a Moving average model of order 1 - ARMA(0,1) can be used 
which is also a ARIMA(0,1,1) with difference 1.

```{r}
# Plotting the partial correlogram for diff1
pacf(death_uk_ts_diff1, lag.max = 20)
pacf(death_uk_ts_diff1, lag.max = 20, plot = FALSE)

```
The partial autocorrelation at lags 1,2,and 3 exceeds the significance bounds.
an Auto regressive model of order 3 is possible. ARIMA(3,1,0)


From the principle of parsimony (fewer is better).


MODEL 2 - MOVING AVERAGE MODEL OF ORDER 1 - ARIMA(0,1,1)
```{r}
# Moving average model of order 1 and difference 1.
death_uk_ts_ma <- arima(death_uk_ts, order = c(0,1,1))
death_uk_ts_ma
```

```{r}
# forecasting the next 10 years using moving average.
death_uk_ts_ma_forecast <- forecast(death_uk_ts_ma, h =10)
death_uk_ts_ma_forecast
```

```{r}
# 10 year forecast plot
plot(death_uk_ts_ma_forecast)
```

```{r}
# Evaluation for ARIMA(0,1,1) 
AIC(death_uk_ts_ma)
BIC(death_uk_ts_ma)
accuracy(death_uk_ts_ma)
```


```{r}
# ACF and Ljung box test
acf(death_uk_ts_ma_forecast$residuals, lag.max=20 , na.action = na.pass) 
Box.test(death_uk_ts_ma_forecast$residuals, lag=20, type="Ljung-Box")
```
The P-value for the LJung-test is 0.052, there is little evidence of non-zero auto correlations in the forecast errors at lags 1-20.

```{r}
#  plotting the forecast errors to check for constant variance
plot.ts(death_uk_ts_ma_forecast$residuals, main = 'ARIMA(0,1,1) Forecast Errors',
        ylab = 'Residuals')
```

Appears to have mean 0 and constant Variance

```{r}
# plotting if the forecast errors to check if normally distributed with mean zero
plotForecastErrors(death_uk_ts_ma_forecast$residuals)
```

Appears normally distributed with mean 0



MODEL 3 - ARIMA(3,1,0)
```{r}
# ARIMA model (3,1,0)
death_uk_ts_ar <- arima(death_uk_ts, order = c(3,1,0))
death_uk_ts_ar
```

```{r}
# forecasting the next 10 years.
death_uk_ts_ar_forecast <- forecast(death_uk_ts_ar, h =10)
death_uk_ts_ar_forecast
```

```{r}
# 10 year forecast plot
plot(death_uk_ts_ar_forecast)
```

```{r}
# Evaluation for ARIMA(3,1,0) 
AIC(death_uk_ts_ar)
BIC(death_uk_ts_ar)
accuracy(death_uk_ts_ar)
```


```{r}
# ACF and Ljung box test
acf(death_uk_ts_ar_forecast$residuals, lag.max=20 , na.action = na.pass) 
Box.test(death_uk_ts_ar_forecast$residuals, lag=20, type="Ljung-Box")
```
The P-value for the LJung-test is 0.057, there is little evidence of non-zero auto correlations in the forecast errors at lags 1-20.


```{r}
# plotting the forecast errors to check for constant variance
plot.ts(death_uk_ts_ar_forecast$residuals,main ='ARIMA(3,1,0) Forecast Errors',
        ylab = 'Residuals')
```

```{r}
# plotting if the forecast errors to check if normally distributed with mean zero
plotForecastErrors(death_uk_ts_ar_forecast$residuals)
```



